---
title: "R Notebook"
output: html_notebook
---
# Generate BRFSS Dataset
Read in entire BRFSS dataset, subsample, and process select columns

Details about BRFSS dataset: https://www.cdc.gov/brfss/annual_data/annual_2016.html

The original XPT dataset must be downloaded from here:
https://www.cdc.gov/brfss/annual_data/2016/files/LLCP2016XPT.zip

Note that this is an optional step.  Processed data is in `dat/BRFSS.48K.csv`, or you can obtain the processed dataset with,
```
data.fn<-"https://raw.githubusercontent.com/kijohnson/Advanced-Data-Analysis/master/Class%204%20Data%20Visualization/BRFSS.48K.csv"
BRFSS<-read.csv(data.fn)
```

###Load the oreign library to access the read.xport function
```{r}
library(foreign)
```

###Read in the BRFSS SAS dataset 

```{r}
BRFSS.fn<-"../dat/LLCP2016.XPT"  # this may need to be adjusted 
BRFSS2016<-read.xport(BRFSS.fn)  
```
###Take a 10% random sample of the BRFSS2016 dataset and list all the variables 
```{r}
BRFSS2016_10Percent<-BRFSS2016[sample(1:nrow(BRFSS2016), 48630, replace=FALSE),]
varlist<-colnames(BRFSS2016_10Percent)
varlist
```

###select variables of interest to make the size manageable for the class demonstration
```{r}
myvars<-c("X_STATE", "EMPLOY1", "INCOME2", "SEATBELT",  "DIABETE3", "X_BMI5", "HTM4", "WTKG3", "SEX", "X_AGE80")
BRFSS_samp<-BRFSS2016_10Percent[myvars]
```

###Look at State variable
```{r}
class(BRFSS_samp$X_STATE) #first find class
table(BRFSS_samp$X_STATE) #See how many in each number
```

###Assign labels to values of state (code book can be accessed here: https://www.cdc.gov/brfss/annual_data/2016/pdf/codebook16_llcp.pdf)
```{r}
BRFSS_samp$state<-factor(BRFSS_samp$X_STATE,
        levels=c(1,2,4,5,6,8,9,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,
          37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56,66,72,78),         
        labels=c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO","MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "GUAM", "PR", "VI"))
table(BRFSS_samp$state)
```
###Quality control check to make sure numbers of subjects for X_State correspond with what is reported here https://www.cdc.gov/brfss/annual_data/2016/pdf/codebook16_llcp.pdf in the X_STATE variable
```{r}
BRFSS2016$state<-factor(BRFSS2016$X_STATE,
        levels=c(1,2,4,5,6,8,9,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,
          37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56,66,72,78),         
        labels=c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "GUAM", "PR", "VI"))
table(BRFSS2016$state)  #numbers cross check
```
###Label other variables that we may use: EMPLOY1, INCOME2, SEATBELT, BMI5, DIABETE3 (code book can be accessed here: https://www.cdc.gov/brfss/annual_data/2016/pdf/codebook16_llcp.pdf)
```{r}

BRFSS_samp$employed <- factor(
  BRFSS_samp$EMPLOY1,
  levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 99),
  labels = c("Employed for wages", "Self-employed", "Out of work for 1 year or more", "Out of work for less than 1 year",
  "A homemaker", "A student","Retired","Unable to work","Refused","Not asked or Missing")
  )
  table(BRFSS_samp$employed)
  
BRFSS_samp$income <- factor(
  BRFSS_samp$INCOME2,
  levels = c(1, 2, 3, 4, 5, 6, 7, 8, 77, 99, "BLANK"),
  labels = c(
  "< $10,000",
  "$10,000 to less than $15,000",
  "$15,000 to less than $20,000",
  "$20,000 to less than $25,000",
  "$25,000 to less than $35,000",
  "$35,000 to less than $50,000",
  "$50,000 to less than $75,000",
  "$75,000 or more", "Don't know/Not sure",
  "Refused","Not asked or missing")
  ) 
  table(BRFSS_samp$income)
  
BRFSS_samp$seatbelt <- factor(
  BRFSS_samp$SEATBELT,
  levels = c(1, 2, 3, 4, 5, 7, 8, 9),
  labels = c(
  "Always",
  "Nearly always",
  "Sometimes",
  "Seldom",
  "Never",
  "Don't know/Not sure",
  "Never drive or ride in a care",
  "Refused")
  )
  table(BRFSS_samp$seatbelt)
  
class(BRFSS_samp$X_BMI5)
  
BRFSS_samp$diabetes <- factor(
  BRFSS_samp$DIABETE3,
  levels = c(1, 2, 3, 4, 7, 9),
  labels = c(
  "Yes",
  "Yes, but female told only during pregnancy",
  "No",
  "No, pre-diabetes or borderline diabetes",
  "Don't know/Not Sure",
  "Refused")
)
BRFSS_samp$diabetes_short_label <- factor(
  BRFSS_samp$DIABETE3,
  levels = c(1, 2, 3, 4, 7, 9),
  labels = c(
  "Yes",
  "Yes pregnancy",
  "No",
  "pre/borderline",
  "Unknown",
  "Refused")
)
#table(BRFSS_samp$diabetes)

BRFSS_samp$sex <- factor(
  BRFSS_samp$SEX,
  levels = c(1, 2, 9),
  labels = c("Male", "Female", "Refused")
  )
  table(BRFSS_samp$employed)

# Define height and weight columns in units of m and kg
BRFSS_samp$height <- BRFSS_samp$HTM4/100
BRFSS_samp$weight <- BRFSS_samp$WTKG3/100
BRFSS_samp$bmi <- BRFSS_samp$X_BMI5/100
BRFSS_samp$age <- BRFSS_samp$X_AGE80   # really just rename it
```
Delete columns which have been converted
Write csv file for Matt for class demo
```{r}
drops <- c("X_STATE", "EMPLOY1", "INCOME2", "SEATBELT", "DIABETE3", "X_BMI5", "HTM4", "WTKG3", "SEX", "X_AGE80")
BRFSS_samp <- BRFSS_samp[, !(names(BRFSS_samp) %in% drops)]
```

```{r}
write.csv(BRFSS_samp, "dat/BRFSS.48K.csv", row.names=FALSE, quote=TRUE)
write.table(BRFSS_samp, "dat/BRFSS.48K.tsv", row.names=FALSE, quote=FALSE, sep="\t")
```

