---
title: "Homework 7"
output: html_document
author: "Matt Wyczalkowski <m.wyczalkowski@wustl.edu>"
---

References:
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html

Load the ggplot2 library
```{r}
library(ggplot2)
```
# Read in BRFSS data
Get rid of all rows which have any missing values (NA)
We add a random column to help with testing
```{r}
data.fn<-"dat/BRFSS.48K.csv"
BRFSS<-read.csv(data.fn)
BRFSS <- BRFSS[rowSums(is.na(BRFSS))==0,]
my.levels=c(
"No",
"Yes",
"pre/borderline",
"Yes pregnancy",
"Refused",
"Unknown"
)
BRFSS$diabetes_short_label <- factor(BRFSS$diabetes_short_label, levels = my.levels)
BRFSS$random = sample(1000, size = nrow(BRFSS), replace = TRUE)  # add a column of random numbers for testing
print(head(BRFSS))
```

# Evaluate significant differences
Based on this example: http://www.r-tutor.com/elementary-statistics/non-parametric-methods/mann-whitney-wilcoxon-test

Is this the best test to use?

## Stats on one pair
For now, comparing "Yes" and "No".  `BRFSS.mw` is a subset of entire dataset with just two factors of diabetes: Yes and No

```{r}
BRFSS.mw = BRFSS[BRFSS$diabetes_short_label %in% c("Yes", "No"),]
print(head(BRFSS.mw))
```

Perform statistics.  Seems to be very significant.
```{r}
results = wilcox.test(weight ~ diabetes_short_label, data=BRFSS.mw)
print(results)
print(results$p.value)
```

Convincing ourselves that calling wilcox test correctly by testing correlation with random number.  There is no correlation.
```{r}
results = wilcox.test(random ~ diabetes_short_label, data=BRFSS.mw)
print(results$p.value)

```

# Plotting of all vs. all diabetes comparison

We want to construct a new data frame which has the significance (p-value) between **pairs** of diabetes status values.  To help us do this, we create a function which subsets BRFSS given any pair of diabetes status values `a` and `b` and performs the wilcox test on these, returning the p.value.

```{r}

get.p.value = function(a, b) {
  if (a==b) {
    return(1)
  }
  BRFSS.mw = BRFSS[BRFSS$diabetes_short_label %in% c(a, b),]
  results = wilcox.test(height ~ diabetes_short_label, data=BRFSS.mw)
  return(results$p.value)
}
```

Now call this function in a double loop.  Note that it would be more "elegant" to use one of the `apply` functions, but this does the trick.  We also evaluate the average weight for each diabetes status - these will define the vertical endpoints for the arcs we draw indicating significance.

```{r}
results = NULL
for (A in levels(BRFSS$diabetes_short_label)) {
  for (B in levels(BRFSS$diabetes_short_label)) {
    p = get.p.value(A,B)
    results = rbind(results, data.frame(A=A, B=B, p.value=p,
                    mA=mean(BRFSS[BRFSS$diabetes_short_label==A,]$weight),
                    mB=mean(BRFSS[BRFSS$diabetes_short_label==B,]$weight)))
  }
}
```

To help with the plotting, we create a column which indicates which status pairs are significant.  Since p values for A,B and B,A are the same and we don't want to plot them twice, we keep only one.  
```{r}
results$is.significant = results$p.value < 0.005
results$A.lt.B = as.numeric(results$A) < as.numeric(results$B)
results$to.plot = results$is.significant & results$A.lt.B
results
```

Finally, we are ready to plot the `results` dataset on top of the previous violin plot (`p`). 
Here's the violin plot:
```{r}
p <- ggplot(data=BRFSS)
p <- p + geom_violin(aes(x=diabetes_short_label, y=weight, fill=diabetes_short_label), color=NA) + guides(fill=FALSE)
p <- p + scale_fill_brewer(palette="Set1", name="Diabetes Status")
p <- p + theme_bw() + xlab("Diabetes Status") + ylab("Weight [kg]")
p <- p + theme(panel.grid.major.x = element_blank())
p
```

Use `geom_curve` to draw arcs between all significant pairs.  Note that we replace the `data` of the above figure with `results`.

```{r}
p2 = p + geom_curve(data=results[results$to.plot,], aes(x=A,y=mA,xend=B, yend=mB), curvature=-0.5)
p2 = p2 + ggtitle("Significance < 0.005 shown")
p2
```

Ideas for additional plots - have the thickness of the line vary according to the degree of significance.